What We Eval in the Shadows

### 2021-08-09 paper/main.tex
Fix author list in the paper heading
methododoly

\documentclass[review,screen,acmsmall]{acmart}
%\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\begin{document}
\title{What We Eval in the Shadows}
\subtitle{A large-scale study of {\sf eval} usage in R programs}

@@

\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
that \eval  is widely used in R code with 8500 call sites in 1238 CRAN packages and 2.2 million dynamic calls.   The 15 most frequent call sites account for 88\% of those. The
\c{match.arg} function is the highest used one with 54\% of all calls. In the
other call sites, they saw two use cases. The most common is the evaluation of
the source code of a promise retrieved by \c{substitute} in a new environment,
\eg as done in the \c{with} function. The other use case is the invocation of a
function whose name or arguments are determined dynamically. For this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository\footnote{\url{https://www.bioconductor.org/}} in addition to CRAN. We also track much more runtime information on the \eval calls during tracing. It enables us to have a more nuanced and accurate picture, and identify more diverse use cases of \eval.

@@

This section explains how we selected our corpus and how we obtained the
reported results.

@@

Our corpus is assembled from three sources: the \emph{Base} libraries bundled
with the language implementation, packages hosted on \emph{CRAN}, and scripts
from \emph{Kaggle}.

\mypara{Base.} The dataset contains \BasePackages libraries performing basic
arithmetics, statistics, and operating system functionalities. These libraries
are bundled with R and executed pervasively. Together they contain
\BaseEvalCallSites call sites to \eval in \BaseFunsWithEvals functions. Some of these
functions provide basic functionalities such as package loading, and thus, there
is hardly any code that does not invoke one of these.

\mypara{CRAN.} The Comprehensive R Archive Network
({\small \url{cran.r-project.org}}) is the largest curated repository of R
packages. It hosts \CranAvailablePackagesRnd packages with 6 new ones submitted
daily~\cite{Ligges2017}. Packages are authored by experienced developers and
abide by well-formedness rules automatically checked on each commit.
Each package comes with sample data. There are three sources of
runnable code in a package: \emph{tests, examples}, and \emph{vignettes} --
respectively, unit tests, code snippets from the documentation, and long-form
use-cases written in RMarkdown. Examples and vignettes can be turned into
scripts by extracting the relevant code. From \CranPackages packages we
extracted \CranRunnableScripts scripts that contain \CranRunnableCode lines of
code (\CranRunnableCodeExamplesRnd in examples, \CranRunnableCodeVignettesRnd in
vignettes and \CranRunnableCodeTestsRnd in tests).

\mypara{Kaggle.} The Kaggle website ({\small \url{kaggle.com}}) is an online
platform for data science. It allows users to submit and compete to solve
problems. Solutions, called \emph{kernels}, are uploaded as scripts or
notebooks. While the quality of code is not uniform, each kernel is runnable,
and maps to ones script. Input data is provided. We obtained \KaggleKernels
kernels. Since Kaggle is not curated, we used SHA-1 hashes to identify and
remove \KaggleDuplicates duplicate entries. The remaining \KaggleUnique kernels
are unique solutions to \KaggleCompetitions competitions. They contain
665K lines of R code. Only \KaggleWithEvals kernels call \eval.

@@

The analysis presented in this has been automated by a pipeline that acquires
packages, extracts scripts, executes them, traces their behavior, and summarizes observations. Figure~\ref{fig:pipeline} shows the main steps along
 with their running time, data size, and  number of elements manipulated.
 Timings are from a cluster of three servers, each with 2.3GHz Intel Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network storage. The pipeline steps are:

@@

%% \begin{figure}[!h]\hspace{-5mm}
%%  \includegraphics[width=.95\linewidth]{pipeline.pdf}
%%  \caption{Pipeline}\label{fig:pipeline}
%%\end{figure}

\medskip
\begin{compactenum}
\item \emph{Download.} Packages are downloaded from CRAN. For scripts, a web
  crawler retrieves code, and the Kaggle command-line tool gets data.
  Installation is complicated by native dependencies, which are not adequately
  documented and thus hard to resolve automatically.
\item \emph{Extract.} Given installed packages, the \genthat tool extracts all
  runnable code snippets and turns each of these into a self-standing
  program~\cite{issta18}. Some Kaggle kernels are already scripts and nothing
  more needs to be done. Kernels released as notebooks are processed by
  \c{knitr} which extracts runnable code. The body of each program is
  instrumented with calls to our dynamic analyzer to ensure that we only record
  calls to \eval from the code of interest and not from bootstrapping or
  execution harness operations.
\item \emph{Trace.} Each program executes using our dynamic analysis tool, a
  heavily instrumented interpreter that captures calls to \eval and many other
  fine-grained runtime events. Packages are run twice, once to capture \eval
  calls originating from package code and a second time to capture calls coming
  from the base libraries. To avoid any interference, each program is run in its
  own process with the GNU R compiler turned off to avoid recording its
  execution.
\item \emph{Analyze.} Finally, the analysis output is merged, cleaned, and summarized
  in a post-processing phase driven by series of R scripts. The summarized data
  is then analyzed in RMarkdown notebooks to gather insights. All figures and
  numbers appearing in the paper are generated automatically. Figures are
  produced in PDF by \c{ggplot2}; numbers are exported as \LaTeX macros.
\end{compactenum}

\noindent The pipeline runs in parallel~\cite{GNUparallel} orchestrated
by a Makefile. Servers have identical environments thanks to docker images with
all dependencies installed.

@@

The dynamic analysis is performed by \rdyntrace, a modified R virtual machine
based on GNU R 4.0.2 that exposes low-level callbacks for a variety of runtime
events~\cite{oopsla19b}. The tracer registers callbacks to all \eval variants
and a few additional functions to assist in locating the origin of \eval
arguments. For example, we taint the results of calls to \c{parse} and
\c{match.call}. We also capture calls to R API for dynamic code loading. Next,
we subscribe to the events related to a variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. One challenge was that R only provides source references for
block surrounded by braces. Thus, a function whose body is \c{eval(x)} would no
usable debug information, while the same expression surrounded by braces would
be fine. This is unfortunately not easily fixed. We extended the dynamic
analysis tool to attach synthetic source code references to all \eval call
sites by traversing ASTs. The tracer is implemented as an R package in 3.2K C++
and 1.3K of R code. For performance reasons, most of the tracing is done in
C++. While in theory, the implementation is relatively straightforward, not so
in practice:
%
\begin{compactitem}[---]

\item The lazy evaluation makes it challenging to analyze function arguments
  while tracing as prematurely forcing a promise might have dangerous
  consequences.

\item While the R interpreter is implemented in C, most core functionality is in
  R. For example, package loading is implemented in R using \eval. This makes it
  hard in the tracer to separate the \eval that are essential to user programs
  from the accidental ones that are products of the way R implements its basic
  operations. This is even more so for the side-effects analysis.

\item In the dynamic analysis, we run all the \emph{runnable} code we obtained
  from CRAN and Kaggle---\ie actual code written by people with highly varying
  expertise in R and programming in general. This exercises a lot of the corner
  cases of the highly underspecified R behavior.

\mypara{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source references. This occurs when \eval
is passed as an argument to a higher-order function or when the \eval call
originates from native code. However, these missing source references account
for a meager \PkgUndefinedRatio of all calls; they are unlikely to affect our
results. Other limitations are that we ignore calls to the native \eval and to
the alternate \c{rlang::tidy\_eval} uses native \eval internally. None of these
limitations should invalidate our conclusions.

%\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% It's dubious whether R programmers would develop with macros though.
% Macros and R
% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.  There are no local macro variables in this implementation though.

>>>>

\documentclass[screen,acmsmall]{acmart}
%\settopmatter{printfolios=false,printccs=false,printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

@@

\begin{document}
\title{What We Eval in the Shadows}
\subtitle{A Large-scale Study of Eval in R programs}

@@

\citet{ecoop12} had a few paragraphs on the usage of \eval in R. They found the
that \eval  is widely used in R code with 8500 call sites in 1238 CRAN packages and 2.2 million dynamic calls.   The 15 most frequent call sites account for 88\% of those. The
\c{match.arg} function is the highest used one with 54\% of all calls. In the
other call sites, they saw two use cases. The most common is the evaluation of
the source code of a promise retrieved by \c{substitute} in a new environment,
\eg as done in the \c{with} function. The other use case is the invocation of a
function whose name or arguments are determined dynamically. For this purpose, R provides \c{do.call} and thus \eval is overkill. Our corpus is larger, with \CranPackages packages, \packageTriggeredpkgs packages with at least one \eval, for a total of \PkgHitEvalCallSites sites and \packageAllcallsRnd dynamic calls, although they also looked at the Bioconductor package repository in addition to CRAN. We also track much more runtime information on the \eval calls during tracing. It enables us to have a more nuanced and accurate picture, and identify more diverse use cases of \eval.

@@

This section describes how the corpus of R program was selected and the analysis
infrastructure.

@@

The corpus is assembled from three sources: the \emph{Base} libraries bundled
with the language implementation, packages hosted on \emph{CRAN}, and scripts
from \emph{Kaggle}. Each of these sources represents a different part of
the user base of the language and thus, potentially, different usage patterns.

\mypara{Base.} The \BasePackages libraries are bundled with R and widely used as
they provide arithmetics, statistics, and operating system functionalities. Some
\BaseFunsWithEvals functions contain \BaseEvalCallSites calls to \eval. The
functions perform such key tasks such as package loading, there is thus hardly
any R program that does not invoke them. These libraries have been written by
experts with a deep understanding of the language and its environment.

\mypara{CRAN.} This dataset consists of packages hosted on the Comprehensive R
Archive Network ({\small \url{cran.r-project.org}}, aka CRAN). There are three
sources of runnable code in a package: unit \emph{tests} for individual
functions, code \emph{examples} automatically extracted from the documentation,
and long-form use-cases called \emph{vignettes}. Package also often have sample
input files. All of these runnable code snippets can be extract into independent
files. We downloaded \CranPackages packages and extracted \CranRunnableScripts
scripts, or \CranRunnableCode lines of code (\CranRunnableCodeExamplesRnd in
examples, \CranRunnableCodeVignettesRnd in vignettes and
\CranRunnableCodeTestsRnd in tests). CRAN packages are authored by experienced
developers.

\mypara{Kaggle.} This dataset contains \KaggleUnique programs from the Kaggle
website ({\small \url{kaggle.com}}), an online platform that allows users to
submit and compete to solve problems. Kaggle contains both scripts and notebooks
as well as input files, but there is some duplication (we used SHA-1 hashes to
remove \KaggleDuplicates duplicates). The dataset has 665K lines of R code and
only \KaggleWithEvals calls \eval. Kaggle scripts are written by end-users with
variable levels of expertise.

@@

We implemented an automated pipeline to acquires packages, extracts scripts,
executes them, traces their behavior, and summarizes observations.
Figure~\ref{fig:pipeline} shows the main steps along with their running time,
data size, and number of elements manipulated. The pipeline steps are:

@@

\begin{enumerate}
\item \emph{Download.} Packages are downloaded from CRAN, for Kaggle a web
  crawler retrieves code, and a command-line tool gets data. Installation is
  complicated by native dependencies which sometimes have to be resolved
  manually.
\item \emph{Extract.} The \genthat tool extracts all runnable code snippets and
  turns each of these into a self-standing program~\cite{issta18}, \c{knitr}
  extracts code from notebooks. Each extracted script is instrumented with calls
  to our dynamic analyzer in a way that avoid recording calls to \eval
  originating from our execution harness.
\item \emph{Trace.} Script run with a modified R interpreter to capture
  execution traces. Each script is run in its own process with the GNU R
  compiler turned off to avoid recording its execution. CRAN scripts run twice,
  once to capture \eval calls originating from package code and a second time to
  capture calls coming from base libraries.
\item \emph{Analyze.} Analysis outputs are merged, cleaned, and summarized.
  RMarkdown notebooks process the summarized data to generate all graphs (PDF
  from \c{ggplot2}) and numbers (exported as \LaTeX macros) appearing in the
  paper.
\end{enumerate}

\noindent The pipeline runs in parallel on three servers, each with 2.3GHz Intel
Xeon 6140 processor with 72 cores and 256GB of RAM, and shared OCFS network
storage. Servers run identical docker images with all dependencies installed.

@@

The tracer that performs dynamic analysis of R scripts is built on top of
\rdyntrace, an extended R 4.0.2 virtual machine that exposes callbacks for
various runtime events~\cite{oopsla19b}. The tracer registers callbacks to all
variants of \eval and a few additional functions to locate the origin of \eval
arguments (it taints the results of calls to \c{parse} and \c{match.call}). It
captures dynamic code loading, as well as variable definition and assignment,
allowing us to record side effects that happen in environments while evaluating
code in \eval. A challenge was that lack of source code references for
expression that are not within block surrounded by braces. This is unfortunately
not easily fixed. We extend our dynamic analysis tool to attach synthetic source
code references to all \eval call sites, but the approach fails in some edge
cases. For performance reasons, the tracer is an R package written in C++ (3.2K
LOC) and R (1.3K LOC). While in theory the implementation should be
straightforward, not so in practice. Lazy evaluation requires delaying the
processing of arguments until (and if) they are forced. Accounting for
side-effects performed in an \eval is complicated by the fact that R is
implemented in a mixture of R and C, and that the language implementation can
call \eval. Lastly, large code bases exercise many edge cases of the highly
underspecified R behavior.

\mypara{Limitations.} Even with the extension described above, there are
\PkgUndefinedRnd \eval calls without source information (\PkgUndefinedRatio of
all \eval calls). This happen when \eval is an argument to a higher-order
function or when it is called from native code. An other limitation is that we
do not record calls to the native \eval. Neither of these limitations should
invalidate our conclusions.

\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage Metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@
